Open source symbolic multibody dynamics software

### 2012-02-02 PetersonGedeHubbard2012.tex
Finished unconstrained state equations, nearly done with permutation calculations.

The complete configuration of the system is assumed to be described by
generalized coordinates $\bm{q}\in\bm{R}^n$, its velocity completely described
by generalized speeds $\bm{u}\in\bm{R}^o$.  It is assumed that there are $l$
holonomic constraints ($f_h : \mathbf{R}^n \times \mathbf{R} \to
\mathbf{R}^l$), and $m$ nonholonomic constraints ($f_{nh} : \mathbf{R}^n \times
\mathbf{R}^o \times \mathbf{R} \to \mathbf{R}^l$). Two functions $\bm{f}_0 :
\mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R} \to \mathbf{R}^n$ and
$\bm{f}_1 : \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R} \to
\mathbf{R}^n$ relate the coordinate time derivatives to the generalized speeds,
and two more functions $\bm{f}_2 : \mathbf{R}^n \times \mathbf{R}^o \times
\mathbf{R}^p \times \mathbf{R} \to \mathbf{R}^n$ and $\bm{f}_3 : \mathbf{R}^n
\times \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R}^p \times \mathbf{R}
\to \mathbf{R}^n$ relate the time derivatives of the speeds $\bm{\dot{u}}$ to
the externally applied inputs $\bm{r}\in\bm{R}^q$, coordinates $\bm{q}$ and
their time derivatives $\bm{\dot{q}}$,  speeds $\bm{u}$, and time $t$. These
assumptions can be summarized by the vector equations
\begin{align}
  \label{eq:holonomic}
  \bm{f}_{h}(\bm{q}, t) &= \bm{0}\\
  \label{eq:nonholonomic}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &= \bm{0}\\
  \label{eq:kindiffs}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t)
    + \bm{f}_{1}(\bm{q}, \bm{u}, t) &= \bm{0} \\
  \label{eq:dyndiffs}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, \bm{r}, t)
    + \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) & = \bm{0}
\end{align}
where (\ref{eq:holonomic}) is the holonomic constraints,
(\ref{eq:nonholonomic}) is the nonholonomic and time-differentiated holonomic
constraints, (\ref{eq:kindiffs}) is the kinematic differential equations
(linear in both $\bm{\dot{q}}$ and $\bm{u}$ and in the trivial case,
$\bm{\dot{q}} = \bm{u}$, so $f_0 = -f_1$), and (\ref{eq:dyndiffs}) include both
the dynamic differential equations (arising from application of Kane's method)
and the time-differentiated version of equation \ref{eq:nonholonomic}.
Equation (\ref{eq:dyndiffs}) will necessarily by linear in $\bm{\dot{u}}$, but
generally nonlinear in all other quantities (though often linear in the
external inputs $\bm{r}$).  In this formalism, both independent and dependent
quantities (coordinates and speeds) appear in these equations; we will address
this is shortly.

@@

We begin with a Taylor series expansion of equations
\ref{eq:holonomic}-\ref{eq:dyndiffs} about an equilibrium $\bm{q}=\bm{q}^*$,
$\bm{\dot{q}}=\bm{\dot{q}}^*$, $\bm{u}=\bm{u}^*$,
$\bm{\dot{u}}=\bm{\dot{u}}^*$.  It assumed that $\bm{q}^*, \bm{\dot{q}}^*,
\bm{u}^*, \bm{\dot{u}}^*$ satisify equations
\ref{eq:holonomic}-\ref{eq:dyndiffs}.  The Taylor series expansion of the
constraint equations (\ref{eq:holonomic}) and (\ref{eq:nonholonomic}) about the
equilibrium is
\begin{align}
  \label{eq:holonomic_expansion}
  \bm{f}_{h}(\bm{q}, t) &\approx \underbrace{f_{h}(\bm{q}^*, t)}_{\bm{0}} +
    \nabla_{\bm{q}}f_{h}(\bm{q}^*, t)\delta \bm{q}\\
  \label{eq:nonholonomic_expansion}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &\approx \underbrace{\bm{f}_{nh}(\bm{q}^*,
  \bm{u}^*, t)}_{\bm{0}} +  \nabla_{\bm{q}}f_{nh}(\bm{q}^*, \bm{u}^*, t)\delta \bm{q} + \nabla_{\bm{u}}f_{nh}(\bm{q}^*, \bm{u}^*, t)\delta \bm{u}
\end{align}
The

>>>>

The complete configuration of the system is assumed to be described by
generalized coordinates $\bm{q}\in\bm{R}^n$, its velocity completely described
by generalized speeds $\bm{u}\in\bm{R}^o$.  It is assumed that there are $l$
holonomic constraints ($f_h : \mathbf{R}^n \times \mathbf{R} \to
\mathbf{R}^l$), and $m$ nonholonomic constraints ($f_{nh} : \mathbf{R}^n \times
\mathbf{R}^o \times \mathbf{R} \to \mathbf{R}^l$). Two functions $\bm{f}_0 :
\mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R} \to \mathbf{R}^n$ and
$\bm{f}_1 : \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R} \to
\mathbf{R}^n$ relate the coordinate time derivatives to the generalized speeds,
and two more functions $\bm{f}_2 : \mathbf{R}^n \times \mathbf{R}^o \times
\mathbf{R} \to \mathbf{R}^o$ and $\bm{f}_3 : \mathbf{R}^n
\times \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R}^p \times \mathbf{R}
\to \mathbf{R}^o$ relate the time derivatives of the speeds $\bm{\dot{u}}$ to
the externally applied inputs $\bm{r}\in\bm{R}^q$, coordinates $\bm{q}$ and
their time derivatives $\bm{\dot{q}}$,  speeds $\bm{u}$, and time $t$. These
assumptions can be summarized by the vector equations
\begin{align}
  \label{eq:holonomic}
  \bm{f}_{h}(\bm{q}, t) &= \bm{0}\\
  \label{eq:nonholonomic}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &= \bm{0}\\
  \label{eq:kindiffs}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t)
    + \bm{f}_{1}(\bm{q}, \bm{u}, t) &= \bm{0} \\
  \label{eq:dyndiffs}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t)
    + \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) & = \bm{0}
\end{align}
where equations (\ref{eq:holonomic}) and (\ref{eq:nonholonomic}) are the
holonomic and nonholonomic constraints (and possibly time differentiated holonomic
constraints), respectively. Equations (\ref{eq:kindiffs}) are the kinematic differential equations
(linear in both $\bm{\dot{q}}$ and $\bm{u}$ and in the trivial case,
$\bm{\dot{q}} = \bm{u}$, so $f_0 = -f_1$).  Equations (\ref{eq:dyndiffs}) are
the dynamic differential equations
and the time differentiated version of equation \ref{eq:nonholonomic}.
Equation (\ref{eq:dyndiffs}) will necessarily by linear in $\bm{\dot{u}}$, but
generally nonlinear in all other quantities (though often linear in the
external inputs $\bm{r}$).  In this formalism, both independent and dependent
quantities (coordinates and speeds) appear in these equations; we will address
this is shortly.  This following calculations formalism is exceedingly generic;
in practice there is a lot of structure in equations
(\ref{eq:nonholonomic}-\ref{eq:dyndiffs}) that can be leveraged to simplify the
following calculations.  This will be expanded upon below.

@@

We begin with a first order Taylor series expansion of equations
(\ref{eq:holonomic})-(\ref{eq:dyndiffs}) about an equilibrium
$\bm{q}=\bm{q}^*$, $\bm{\dot{q}}=\bm{\dot{q}}^*$, $\bm{u}=\bm{u}^*$,
$\bm{\dot{u}}=\bm{\dot{u}}^*$, $\bm{r}=\bm{r}^*$ that satisfies equations
(\ref{eq:holonomic})-(\ref{eq:dyndiffs}).  Determining these equilibrium values
generally involves solving (\ref{eq:holonomic})-(\ref{eq:dyndiffs}).  In the interest of brevity, we omit writing this equilibrium after
each gradient in the calculations below; all are evaluated at these equilibrium
conditions.  Taylor series expansion of the constraint equations
(\ref{eq:holonomic}) and (\ref{eq:nonholonomic}) about the equilibrium is
\begin{align}
  \label{eq:holonomic_expansion}
  \bm{f}_{h}(\bm{q}, t) &\approx \underbrace{f_{h}(\bm{q}^*, t)}_{\bm{0}} +
    \nabla_{\bm{q}}f_{h} \cdot \delta \bm{q}\\
  \label{eq:nonholonomic_expansion}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &\approx \underbrace{\bm{f}_{nh}(\bm{q}^*,
  \bm{u}^*, t)}_{\bm{0}} +  \nabla_{\bm{q}}f_{nh} \cdot \delta \bm{q} +
  \nabla_{\bm{u}}f_{nh} \cdot \delta \bm{u}
\end{align}
and the first terms are identically zero because of the assumption that the
equilibrium point satisfies the constraints.  The Taylor series expansions of
the functions appearing in the left hand side of equations (\ref{eq:kindiffs})
are
\begin{align}
  \label{eq:f0_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) &\approx \bm{f}_{0}(\bm{q}^*,
  \bm{\dot{q}}^*, t) + \nabla_{\bm{q}}\bm{f}_{0} \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}}\\
  \label{eq:f1_expansion}
  \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx \bm{f}_{1}(\bm{q}^*,
  \bm{u}^*, t) + \nabla_{\bm{q}}\bm{f}_{1} \cdot \delta\bm{q} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Summing (\ref{eq:f0_expansion}) and (\ref{eq:f1_expansion}) and recognizing
that the sum of the first term on the right hand side of each equation must
equal zero, we obtain
\begin{align}
  \label{eq:f0_plus_f1_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx
  \nabla_{\bm{q}}(\bm{f}_{0} + \bm{f}_{1}) \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Similarly, a Taylor series expansion of the functions appearing in
the left hand side of (\ref{eq:dyndiffs}) yields
\begin{align}
  \label{eq:f2_expansion}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) &\approx
      \bm{f}_{2}(\bm{q}^*, \bm{\dot{u}}^*, t) +
      \nabla_{\bm{q}}\bm{f}_{2} \cdot \delta\bm{q}
      + \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}}\\
  \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) &\approx
  \bm{f}_{3}(\bm{q}^*, \bm{\dot{q}}^*, \bm{u}^*, \bm{r}^*, t) +
  \nabla_{\bm{q}}\bm{f}_{3} \cdot \delta\bm{q}\notag\\
  \label{eq:f3_expansion}
  &+ \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}
  + \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta \bm{u}
  + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}
Summing (\ref{eq:f2_expansion}) and (\ref{eq:f3_expansion}) and recognizing
that the sum of the first term on the right hand sides of these equations must
equal zero, we obtain
\begin{align}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}},
  \bm{u}, \bm{r}, t) &\approx \nabla_{\bm{q}}(\bm{f}_2 + \bm{f}_3) \cdot
  \delta\bm{q} + \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}\notag\\
  \label{eq:f2_plus_f3_expansion}
  &+ \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta\bm{u} +
  \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}} + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}

Utilizing equations (\ref{eq:kindiffs}), (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:dyndiffs}), and (\ref{eq:f2_plus_f3_expansion}), we introduce the
following quantities
\begin{align}
  M_{00} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 \\
  M_{11} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 \\
  A_{00} &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
  A_{01} &\triangleq -\nabla_{\bm{u}}\bm{f}_1 \\
  A_{10} &\triangleq \nabla_{\bm{q}}(\bm{f}_2 + \bm{f}_3)
  -\nabla_{\bm{\dot{q}}}\bm{f}_3 (\nabla_{\bm{\dot{q}}}\bm{f}_{0})^{-1}
  \nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
  A_{11} &\triangleq \nabla_{\bm{u}} \bm{f}_{3} - \nabla_{\bm{\dot{q}}}\bm{f}_{3}
  (\nabla_{\bm{\dot{q}}}\bm{f}_0)^{-1} \nabla_{\bm{u}}\bm{f}_{1} \\
  B_{1} &\triangleq \nabla_{\bm{r}}\bm{f}_{3}
\end{align}

Equating the right hand side of equations (\ref{eq:f0_plus_f1_expansion}) and
(\ref{eq:f2_plus_f3_expansion}) each to zero, eliminating $\bm{\dot{q}}$ from
(\ref{eq:f2_plus_f3_expansion}), and rewriting in matrix form, we obtain
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      M_{00} & \bm{0}_{n \times o} \\
      \bm{0}_{o \times n} & M_{11}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       A_{00} & A_{01} \\
       A_{10} & A_{11}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{n \times q} \\
        B_{1}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet as was discussed earlier, only $p = n - l + o - m$ of these
quantities are independent.  To address this issue, one must pick a particular
set of independent coordinates and speeds and appeal to equations
(\ref{eq:holonomic_expansion}) and (\ref{eq:nonholonomic_expansion}).  To this
end, consider the following partitioning of the generalized coordinates and
generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.

Setting the right hand side of equation (\ref{eq:holonomic_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{h} P_{q} \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{h} P_{qi} \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{h} P_{qd} \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{h} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{h} P_{qi}) \delta \bm{q}_i
\end{align}
